### 2021-06-03 text/article_vedrovice.Rmd
update text after proofreading

---
title           : "Spatial Analysis and Grave Goods Distribution at the Cemetery of Vedrovice"
# "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"

author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "Spatial patterns in cemeteries and grave goods associations at the
  cemetery of Vedrovice - Široká u Lesa are explored using computationaly
  instensive formal methods.
  Resampling approaches both to intrasite spatial arrangement and
  distribution of grave goods are explored in
  relation to layers of identity represented by sex, age and origin of the
  buried bodies.
  As a result of the investigation we conclude that...  "

output:
    bookdown::pdf_document2:
    # bookdown::word_document2:
        latex: pdflatex
        toc: false
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "european-journal-of-archaeology.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

The Cemetery of Vedrovice is one of the earliest and most intensively studied
*Linearbandkeramik* (LBK) cemeteries.
Recent contributions use the evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020] and
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
Here we focus on exploration of relationship between three proposed *layers* of
identity, biological sex and age of the body determined from sekeletal remains
by anthropological methods and origin of the body based on Strontium isotope
signal.
We explore whether the structure observed in these three categories is
mirrored in the grave goods associations and spatial organization within the
cemetery.
The analysis is set into a broader discussion of gender relations and identity
inside Neolithic communities.

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
Authors present gender in the Neolihic as being
of a very different nature than gender in the later prehistory.
Gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

<!-- We focus on relations between burials and bodies in terms of biological
sex, age and origin, exceptionality of grave goods and location within the
cemetery.-->
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *layers* of Neolithic identity, i.e. sex, age and localness of the body.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods distributions and spatial
arrangement of the cemetery.

@@

<!-- resampling and computer-intensive techniques-->
The approach taken is a quantitative one embracing resampling techniques and
computer-intensive methods in general.
The resampling techniques were introduced to mortuary studies by
@manly1996 but the application is still limited [@sosna2008, 343].
As noted by Sosna (ibid.), the most important benefit of resampling is in its
ability to distinguish random and non-random patterns.
This is crucial for mortuary analysis because the samples studied, including
the one presented here, are usually small in size and identifying which
observed patterns are caused by mere chance associated with small numbers
and which are actually present.

<!-- ### Questions
* Is there evidence that certain artefacts covary,
either positively or negatively?
Do certain artefacts tend to occur together, or are certain combinations
absent or rare?
* How can the presence and absence data best display which burials
contain similar artefacts,
and which artefacts have similar distributions over burials?
* *How can the data be analysed so as to show most clearly the relationships*
*between the different* **burials** in terms of their **contents**?
* *How can the data be analysed so as to show most clearly the relationship*
*between the different* **artefacts** in terms of their
**distributions in graves**?
* How are grave good assemblages related to age and sex?
(... variations in burial rite coinciding with age or sex of the deceased ...) -->

@@

<!-- site -->
Vedrovice site is situated in the south of Moravia (Czech Republic).
Earliest known finds and archaeological explorations here are from
the 19th century but the excavation took place in the second half of 20th century.
There are two known burial areas, *Za Dvorem* and larger one *Široká u Lesa* and
a settlement area.
For more details on the site refer to @masclanslatorre2020 or the
original publication [@podborsky2002].
The material analyzed here comes from Vedrovice *Široká u lesa* cemetery.

<!-- sample (burials, artefacts) -->
Originally, this burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities [@podborsky2002].
These burials are excluded from most of the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved LBK burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age of the bodies were corrected based on
analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was adapted from @hrncir2020.

Each burial was originally characterized by 20 variables recording counts of
artefacts present in the graves.
Some of  these original variables are omitted in the analysis due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being of no use in establishing relationships between pairs of burials.
Only grave goods occuring in more than 5% of the graves are used in the analysis,
this step reduces the count of the variables to 14.
These include variables noting quantities of various types of pottery vessels in
the graves, counts of chipped stone tools made of either local or non-local raw materials,
presence of polished and beveled stone adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornament.
The personal ornament includes beads made of spondylus shell,
spondylus pendants or oval shaped buckles.
Last but not least presence of ochre pigment in the grave is also noted.
Quantities of various *grave goods* are simplified into a binary matrix indicating
presence or absence of a given trait.

Another category of variables is derived from retrieved human skeletal remains.
These include biological sex, age simplified into age groups and indicator
variable whether the Strontium signal in the remains is of local origin.
The variable representing the recorded sex is simplified into the following
categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
non adult (*n. a.*) and indetermined sex (*ind.*).
Age of the body is also simplified into three categories: juveniles (*juv.*)
comprising of groups Infans and Juvenis, adults (*ad.*, all bodies aged as Adultus)
and mature bodies (*mat.*, anthropological categories Adultus-Maturus, Maturus and
Senilis) and bodies of indetermined age (*ind.*).
Last category is the indicator for the origin of the buried person.
Local Strontium range is defined as two standard deviations from mean of Sr ranges of
bodies with age category determined as Infans.
Bodies faling within this range are labeled as local, outside the range as non-local
and where data on Sr values are missing origin is not determined (*ind.*).

We consider there three variables, sex, age and origin to be indicative of
ones identity in the Neolithic.
To explore the relations between the variables we use them both as individual
categories and combined into following indicator variables -- (1) sex/age,
(2) origin/sex, (3) origin/age and finally (4) origin/sex/age.
Due to frequent missing or indetermined values we decided to limit the studied
sample only to bodies with all of the categories determined and only categories
that occure in more than 5% of burials.

@@

## Randomness of artefact combinations

@@

To determine non-randomness of various artefact combinations in burials we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original co-occurence matrix is permutated *n = 9999* times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifact across the burials are fixed.

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The *S* statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the *v* statistic is a measure of difference between co-occurences of
individual artefact with other artefacts
[@manly1996, 475, for details see SI].

To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.

@@

Spatial patterning in prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as its centroid location in a two dimensional space.
A polygon of idealized excavation extent is used as a study area (a window).
<!-- Using a window closely approximating locations of the burials was considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the borders and empty areas
of the original cemetery. -->

Overall structure of the cemeter is examined using
<!-- either density based approaches or -->
methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
<!-- Density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence its location.-->
These methods are based on distances between pairs and triples of neighbours and
characterize the point pattern based on *second- and third-order properties*
[@nakoinz2016, 130-135; @osullivan2010, 107].
These are in contrast to the assumption underlying *first-order effects* that
any point location is independent on other point locations and
defferent parameters influence its location [@osullivan2010, 126].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
In case of a grave placement in a burial ground we presume that the choice of
a graves location is either random or influenced predominantly by other (previous)
grave locations and
is largely independent on the outside parameters.
In other words *first-order properties* are of little to no importance in
our point of view.
The *second-order properties* and possibly *third-order properties*
are thus considered major factors in locating of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

<!-- To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution). -->
We utilize *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] functions.
These functions are based on nearest neighbour distances and are commonly used
to characterize point patterns in archaeology [@nakoinz2016, 135-144].
The *G* function is a metric of how close events are and summarises
cummulative frequency distribution of the nearest neighbour distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area (window).
The *F* function, an empty-space function, is based on simulating a point pattern
and calculating of distances between random locations and the observed locations.
It is a measure of how far observed events are from any arbitrary location [@osullivan2010, 132-135].
Ripley's *K* function is in comparison with the *G* function resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
We find it the most useful in most of archaeology use cases, cf. [@negre2018].
Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].
For formulas and details on individual functions see the SI.

Given statistics for independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although smaller number of permutations should be sufficient to achieve
credible results
[@baddeley2016, 384] we follow the approach taken by @sayer2013 [78].
Ten most extreme realizations of the simulation are discarded as outliers so
the significance level is 0.01 [@baddeley2016 410].

To explore frequency of neighboring burials falling into same categories we
explore several methods.
A modification of *K* function for multitype point patterns
does not take into account all of the
neighboring individuals in the point pattern but pairs of observations of the
same value within the given category are explored.
Same as *K* function is indicative of the distances at which clustering or
repulsive processes take place, the multitype *K* function shows these distances
for given pairs of levels within a given categorical variable.
The envelopes are constructed using the same procedure as described for other
functions.

To further establish the neighborhood structure we construct neighborhood graphs.
We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
The Delaunay triangulation is considered as a maximum number of neighbors per individual case
scenario while the Gabriel graph the opposite.
For each node, mean number of neighbors of a given category is determined and recorded.
To check how common the observed value is, a permutation test is performed
where the categories are randomly assigned while preserving the overall
proportion of given categories and the neighborhood structure determined by the
graph.
The observed values of neighbor categories are compared against distributions
derived from 999 permutations of the simulated data.

<!-- Percolation analysis -->
Another approach we consider for establishing neigborhood relations is by
constructing buffer zones around each of the burials.
To avoid determinism in designating the radius of the buffer zone around each of
the individuals we utilize percolation analysis.
Percolation analysis is a data centric method used to identify clusters of points based
on spatial separation measured by Euclidean distance [@maddison2020, 269].
The introduction of percolation analysis to archaeology is relatively recent,
nevertheless we consider its effectiveness and usefulness in studies with varied scales
well proven [@maddison2020].
Percolation analysis is based on city clustering algorithm [@rozenfeld2008] and
its generalization employing Euclidean distance [@rozenfeld2011].
Nodes being in a given distance threshold (radius) are assigned into a cluster.
This process is repeated for a set of thresholds while recording maximum
cluster sizes and possibly other statistics.
This allows to identify distances at which clusters exhibit rapid growth (absorbing
large number of nodes) and distances at which the clusters remain relatively stable.
For more details refer to @maddison2020 and associated R package [@schmidt2020].

By means of percolation analysis we identify threshold distances at which cluster
sizes are growing and clusters formed at these distances are further explored.
For each cluster number of individual bodies of a given category is recorded.
The same procedure is repeated for 999 resamples with the given
categories assigned at random with constant probabilities.
Observed numbers of occurences deviating by two standard deviations from the mean of the
resampled results are regarded as strong evidence of non-random patterns being
present in given clusters.
To achieve stability of results and eliminate observations occuring due to
chance in small clusters with low numbers of categories of interest,
the whole resampling procedure is repeated three times and only observations present in
all the iterations are reported with expected numbers of occurence and their
deviations from observed values summarized as mean values.

@@

<!-- software packages -->
Data is analyzed in an R environment for statistical computing and graphics
[@rcore] using functions of several different packages, for the complete list
see the supplementary information.

@@

## Artefact combinations

By counting co-occurences of analyzed variables on simulated data matrices and comparing
the distributions of the statistics proposed by @manly1996 with the values calculated for
the original dataset we can establish whether the overall pattern of artefacts
distribution in the graves is non-random and further identify artefacts that deviate
from the random pattern.
The overall deviation from expected co-occurences, i.e. the *S* statistic
[@manly1996, 475] is significantly large at the 0.05 alpha level.
This suggests that there is strong evidence that the overall distribution of
artefacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->

Examining distributions of individual artefacts we find that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:cooc-vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and buckles.

The relationships between artefact types frequently occuring and missing together
in graves are visualized in Fig. (\@ref(fig:cooc-nets)) with
clusters established using fast-greedy algorithm [@newman2004] and strength of the
given association expressed by the thickness of the edge connecting the nodes.
Artifacts most frequently present with other artifacts are with 5 links grinding tools and
ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
spondylus beads.
The association is strongest between spondylus beads and (a) ceramic bottles and
(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
Artifacts that are frequently absent in the graves together are with 6 links
local lithic tools, globular pots and ceramic bowls and spondylus buckels with
5 and 4 links respectively.
The strongest antagonist relationships are between globular pots and (a) bowls and
(b) adzes.

@@

# Conclusions

Our goal was to identify non-random differences between burials both in terms
of the artefact contents of the graves and position within the cemetery.
The burials or more precisely the buried bodies were divided into categories
based on three variables commonly connected to the identity in the Neolithic period.
These categories are biological sex, age and origin.
The exploration was done both for the categories as single variables and for
all of their combinations.

<!-- Artefact combinations -->

>>>>

---
title           : "Spatial Patterns and Grave Goods Differences at the Cemetery of Vedrovice (Czech Republic): A Resampling Approach to Identity Markers in the Early Neolithic"
author          : "Petr Pajdla"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "The spatial pattern and grave good associations at the cemetery of Vedrovice - Široká u Lesa are explored using computationally intensive formal methods. Resampling approaches to both intra-site spatial arrangement and the distribution of grave goods are explored in relation to layers of identity represented by different buried body attributes – sex, age, geographical origin and their combinations. We found that polished stone adzes, grinding tools, ceramic bowls and spondylus beads occur among non-randomly distributed grave goods. As a result of the investigation of differences among the non-randomly distributed artifacts and attributes of burials, we conclude that the most significant are dichotomies between genders, accentuated by both origin and age. Spatial analysis shows clear trends towards clustering between burials, and the emerging clusters are further explored for similarities in burial attributes."
output:
    bookdown::word_document2:
        latex: pdflatex
        toc: false
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "apa.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

The Cemetery of Vedrovice is one of the earliest and most intensively studied
*Linearbandkeramik* (LBK) cemeteries.
Recent articles use evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020],
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
Here we focus on an exploration of the relationship between three proposed *layers* of
identity, biological sex and age of the body determined from skeletal remains
by anthropological methods and the origin of the body based on the strontium isotope
signal.
We explore whether the structure observed in these three categories is
mirrored in the grave good associations and spatial organization within the
cemetery.
The analysis is set into a broader discussion of gender relationships and identity
in Neolithic communities.

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [p. 128] noted a lack of debate in archaeology on gender for the
European Neolithic compared to other periods.
The authors present gender in the Neolithic as being
of a very different nature than gender in later prehistory.
It is argued that gender is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories [@robb2017, p. 140--142].
Considering the Neolithic burial evidence across Europe, authors find that it
"shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness" [@robb2017, p. 138].
In response, @bickle2020 [p. 14] argues that "biological sex may not have defined gender
during the LBK" although the "sexed bodies did matter".

<!-- We focus on relations between burials and bodies in terms of biological
sex, age and origin, exceptionality of grave goods and location within the
cemetery.-->
We focus mainly on investigating non-random patterns in artifact
co-occurrences and spatial arrangement in the cemetery in relation to the
proposed *layers* of Neolithic identity, i.e., sex, age and localness of the body.
Our question is thus how grave goods relate to the aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity* and spatial
arrangement of the cemetery.

@@

<!-- resampling and computer-intensive techniques-->
The chosen approach is a quantitative one embracing resampling techniques and
computer-intensive methods in general.
The resampling techniques were introduced in mortuary studies by
@manly1996 but the application is still limited [@sosna2008, p. 343].
As noted by @sosna2008, the most important benefit of resampling is its
ability to distinguish random and non-random patterns.
This is crucial for mortuary analysis because the studied samples, including
the one presented here, are usually small in size and identifying which
observed patterns are caused by mere chance associated with small numbers
and which are actually present is problematic.

<!-- ### Questions
* Is there evidence that certain artifacts covary,
either positively or negatively?
Do certain artifacts tend to occur together, or are certain combinations
absent or rare?
* How can the presence and absence data best display which burials
contain similar artifacts,
and which artifacts have similar distributions over burials?
* *How can the data be analyzed so as to show most clearly the relationships*
*between the different* **burials** in terms of their **contents**?
* *How can the data be analyzed so as to show most clearly the relationship*
*between the different* **artifacts** in terms of their
**distributions in graves**?
* How are grave good assemblages related to age and sex?
(... variations in burial rite coinciding with age or sex of the deceased ...) -->

@@

<!-- site -->
The Vedrovice site is situated in south Moravia (Czech Republic).
The earliest known finds and archaeological explorations here are from
the 19th century, but excavations were conducted in the second half of 20th century.
There are two known burial areas, *Za Dvorem* and the larger *Široká u Lesa* and
a settlement area.
For more details on the site, see @masclanslatorre2020 or the
original publication by @podborsky2002.
The material analyzed here comes from the *Široká u lesa* cemetery in Vedrovice.

<!-- sample (burials, artifacts) -->
Originally, this burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century and later
agricultural activities.
These burials are excluded from most of the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved LBK burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
and the anthropological determination of the sex and age of the bodies was corrected based on an
analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on the use and function of the artifacts were gathered from
@masclanslatorre2020.
Information on the localness of various bodies was adapted from @hrncir2020.

Each burial was originally characterized by 20 variables recording the number of
artifacts present in the graves.
Some of  these original variables are omitted in the analysis due to their rare occurrence;
for instance, a spondylus arm ring is present in only a single burial,
rendering it useless in establishing relationships between pairs of burials.
Only grave goods occurring in more than 5% of the graves are used in the analysis;
this step reduces the number of variables to 14.
These include variables noting quantities of various types of pottery vessels in
the graves, counts of chipped stone tools made of either local or non-local raw materials,
the presence of polished and beveled stone adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornaments.
The personal ornaments include beads made of spondylus shells,
spondylus pendants and oval shaped buckles.
Last but not least, the presence of ocher pigment in the grave is also noted.
Quantities of various *grave goods* are simplified into a binary matrix indicating
the presence or absence of a given trait.

Another category of variables is derived from retrieved human skeletal remains.
These include biological sex, age simplified into age groups and an indicator
variable as to whether the strontium signal in the remains is of local origin.
The variable representing the recorded sex is simplified into the following
categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
non-adult (*n. a.*) and undetermined sex (*ind.*).
The age of the body is also simplified into three categories: juveniles (*juv.*)
comprised of the groups *Infans* and *Juvenis*, adults (*ad.*, all bodies aged as *Adultus*)
and mature bodies (*mat.*, anthropological categories *Adultus-Maturus*, *Maturus* and
*Senilis*) and bodies of an undetermined age (*ind.*).
The last category is the indicator for the origin of the buried person.
The local strontium range is defined as two standard deviations from the mean of Sr ranges of
bodies with an age category determined as *Infans*.
Bodies falling within this range are labeled as local, outside the range as non-local;
where data on Sr values is missing, origin is not determined (*ind.*).

As mentioned above, we consider these three variables -- sex, age and origin -- to be indicative of
an individual's identity in the Neolithic.
To explore the relationships between the variables, we use them both as individual
categories and combined into the following indicator variables -- (1) sex/age;
(2) origin/sex; (3) origin/age; and (4) origin/sex/age.
Due to frequent missing or undetermined values, we decided to limit the studied
sample only to bodies with all of the categories determined (and only categories
that occur in more than 5% of burials).

@@

## Randomness of artifact combinations

@@

To determine the non-randomness of various artifact combinations in burials, we conducted a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was
introduced into archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
The original co-occurrence matrix was permuted 9,999 times with constant
row and column totals, i.e., both the count of artifacts in individual graves
and the sum of the given type of artifact across the burials are fixed.

To examine which associations of variables, i.e., artifacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance,
the co-occurrences of observed and simulated data
are summarized using two statistics suggested by @manly1996 [475].
The *S* statistic is a measure of overall deviation between the expected and observed
number of artifact occurrences, while
the *v* statistic is a measure of difference between co-occurrences of
individual artifacts with other artifacts.

To assess whether differences between artifacts that are non-randomly distributed
are reflected in relationships between groups based on sex, age, origin and their combinations,
we calculate mean Euclidean distance (ED) between pairs of burials in multidimensional
space settings defined by the non-randomly distributed artifacts.
The observed mean ED is compared to the mean ED distribution  derived by random
reshuffling of the category labels between the burials 9,999 times with the constant
probabilities of appearance of the given category label.
<!-- To infer whether the observed and randomly created mean ED
distributions are samples of the same population we use non-parametric
Kruskal-Wallis test [@kruskal1952].-->
For each of the categories, the deviation of the mean ED from its randomly generated
distributions is expressed as the fraction of random values of ED larger than
the one observed.
The deviation is deemed significant if less than 5 % of
random values are larger than the observed value.
The process is a variation of the one described for resampling in the case of
artifact combinations, and a similar approach is taken by @sosna2008 [346-347].

<!--
To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.-->

@@

Spatial patterning in prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as its centroid location in two dimensional space.
A polygon of the idealized excavation extent is used as the study area (a window).
<!-- Using a window closely approximating locations of the burials was considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the borders and empty areas
of the original cemetery. -->

<!-- Density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence its location.-->
The overall structure of the cemetery is examined using
methods utilizing distances between neighboring burials.<!--[@osullivan2010, 121-154]-->
These methods are based on distances between pairs and threesomes of neighbors and
characterize the point pattern based on *second- and third-order properties*
[@nakoinz2016, 130-135; @osullivan2010, 107].
These are in contrast to the assumption underlying *first-order effects* that
any point location is independent of other point locations and
different parameters influence its location [@osullivan2010, 126].
In the spatial analysis of settlement patterns, individual locations of
settlements (points) can depend on a number of parameters, e.g.,
elevation, soil type, etc.
In the case of a grave placement in a burial ground, we presume that the choice of
the grave location is either random or influenced predominantly by other (previous)
grave locations and
is largely independent of outside parameters.
In other words, *first-order properties* are of little to no importance in
our point of view.
The *second-order properties* and possibly *third-order properties*
are thus considered major factors in the location of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

<!-- To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution). -->
We utilize *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] functions.
These functions are based on the nearest neighbor distances and are commonly used
to characterize point patterns in archaeology [@nakoinz2016, 135-144].
The *G* function is a metric of how close events are and summarizes
cumulative frequency distribution of the nearest neighbor distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area (window).
The *F* function, an empty-space function, is based on simulating a point pattern
and calculating the distances between random locations and the observed locations.
It is a measure of how far observed events are from any arbitrary location [@osullivan2010, 132-135].
Compared to the *G* function, Ripley's *K* function is resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbors [@nakoinz2016, 138].
We find it the most useful in most archaeology cases, cf. @negre2018.
Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].

The given statistics for the independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although a smaller number of permutations should be sufficient to achieve
credible results
[@baddeley2016, 384], we follow the approach taken by @sayer2013 [78].
The ten most extreme realizations of the simulation are discarded as outliers, so
the significance level is 0.01 [cf. @baddeley2016 410].
The remaining realizations of given point processes are used to construct envelopes
around function curves derived from observed data.

To explore the frequency of neighboring burials falling into the same categories, we
explore several methods.
A modification of the *K* function for a multitype point pattern
does not take into account all of the
neighboring individuals in the point pattern but pairs of observations of the
same value within the given category are explored.
In the same way that the *K* function is indicative of the distances at which clustering or
repulsive processes take place, the multitype *K* function shows these distances
for the given pairs of levels within a given categorical variable.
The envelopes are constructed using the same procedure as described for other
functions.

To further explore the neighborhood structure, we construct neighborhood graphs.
We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
The Delaunay triangulation is regarded as the maximum number of neighbors per individual case
scenario, while the Gabriel graph is the opposite.
For each node, the mean number of neighbors of a given category is determined and recorded.
To check how common the observed value is, a permutation test is performed
where the categories are randomly assigned while preserving the overall
proportion of given categories and the neighborhood structure determined by the
graph.
The observed values of neighbor categories are compared to distributions
derived from 999 permutations of the simulated data.

<!-- Percolation analysis -->
Another approach we consider for establishing neighborhood relationships is
constructing buffer zones around each of the burials.
To avoid determinism in designating the radius of the buffer zone around each of
the individuals, we utilize percolation analysis,
a data centric method used to identify clusters of points based
on spatial separation measured by Euclidean distance [@maddison2020, 269].
Although the introduction of percolation analysis in archaeology is relatively recent,
we consider its effectiveness and usefulness in studies with varied scales
well-proven [@maddison2020].
Percolation analysis is based on a city clustering algorithm [@rozenfeld2008] and
its generalization employing Euclidean distance [@rozenfeld2011].
Points falling within a given distance threshold (radius) are added to a cluster.
This process is repeated for a set of thresholds while recording maximum
cluster sizes and possibly other statistics.
This makes it possible to identify distances at which clusters exhibit rapid growth (absorbing
a large number of nodes) and distances at which the clusters remain relatively stable.
For more details, refer to @maddison2020 and the associated R package [@schmidt2020].

By means of percolation analysis, we identify threshold distances at which maximum cluster
size is growing, and clusters formed at these distances are further explored.
For each cluster, the number of individual bodies of a given category is recorded.
The same procedure is repeated for 999 resamples with the given
categories assigned at random with constant probabilities.
The observed numbers of occurrences deviating by more than two standard deviations from the mean of the
resampled results are regarded as strong evidence of non-random patterns being
present in given clusters.
To achieve the stability of results and eliminate observations occurring by
chance in small clusters with low numbers of categories of interest,
the whole resampling procedure is repeated three times and only observations present in
all the iterations are reported with the expected numbers of occurrence and their
deviations from observed values summarized as mean values.

@@

<!-- software packages -->
Data is analyzed in an R environment for statistical computing and graphics
[@rcoreteam2021] using functions of several different packages; for the complete list,
see the supplementary information.

@@

## Artifact combinations

By counting the co-occurrences of analyzed variables on simulated data matrices and comparing
the distributions of the statistics proposed by @manly1996 with the values calculated for
the original data set, we can establish whether the overall pattern of artifact
distribution in the graves is non-random and further identify artifacts that deviate
from the random pattern.
The overall deviation from expected co-occurrences, i.e., the *S* statistic
[@manly1996, 475] is significantly large at the 0.05 alpha level.
This suggests that there is strong evidence that the overall distribution of
artifacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->

Examining distributions of individual artifacts, we find that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed non-randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:cooc-vstat)).
Among the grave goods found to be randomly distributed in the graves
across the burial ground are ocher pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes, spondylus pendants and buckles.

The relationships between artifact types frequently occurring and missing together
in graves are visualized in Fig. \@ref(fig:cooc-nets) with
clusters established using a fast-greedy algorithm [@newman2004] and the strength of the
given association expressed by the thickness of the edge connecting the nodes.
Artifacts most frequently present with other artifacts are grinding tools and
ceramic bowls with five links and adzes, non-local lithic tools, ocher pigment and
spondylus beads with four links.
The association is strongest between spondylus beads and (a) ceramic bottles and
(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
Artifacts frequently absent with other artifacts in the graves are  
globular pots, local lithic tools, ceramic bowls and spondylus buckles
with seven to four links respectively.
The strongest antagonist relationships are between globular pots and (a) bowls and
(b) adzes.

Artifacts non-randomly distributed across the cemetery are further used to
check whether significant differences exist between their occurrences in
different groups defined based on origin, sex, age and their combinations.
Exploring single attributes, the only differences are in the categorization to
differently sexed bodies; there are significant differences
between male- and female-sexed bodies.
In the case of age or origin, no significant differences are observed.
If we combine sex and age categories, the distinction between juvenile, adult and mature
female-sexed bodies and adult male-sexed bodies is evident.
Another observed dichotomy is present between adult males and juveniles (non-adults).
The same relationships are essentially present in the case of combinations of sex and origin attributes.
Local male-sexed bodies have significantly different combinations of non-randomly
distributed artifacts from both local and non-local female-sexed bodies and non-adults.
If all the categories are combined the divisions are even more evident.
Local male-sexed bodies in the *Adultus* age category deviate in the artifact distributions
from adult local and non-local females, *Maturus/Senilis* non-local female-sexed bodies
and juveniles.

@@

# Conclusions and discussion

Our goal was to identify non-random differences between burials both in terms
of the artifact contents of the graves and their position within the cemetery.
The burials or, more precisely, the buried bodies were divided into categories
based on three variables commonly connected to identity in the Neolithic period.
These categories are biological sex, age and origin.
The exploration was conducted both for the categories as single variables and for
all of their combinations.
